---
title: "Testing the impact of volcanic eruption frequency on multidecadal temperature variability"
output: html_notebook
author: Kira Rehfeld 
date: 22.02.2019
---

Reading in the CMIP5 forcing data, translating it to effective radiative forcing, run 0D EBM with different combinations of forcing.  

**To Do/ Questions**

* what is the correct anomaly period/value (forcing)?
* translation AOD to W/m^2 (-25W/m^2 per 1 unit AOD), AR5/AR6 value 


```{r Packages and Functions, message=FALSE, warning=FALSE, paged.print=TRUE}
#rm(list=ls())
library(deSolve)
library(zoo)
library(multitaper)
library(PaleoSpec)
source("ebm_functions.R") 
source("var_from_spec.R")

outputdir<-"~/R/spielwiese/ebm0/plots/"
```
```{r Read Forcing Data}
Ftab<-read.csv("~/mydata/CMIP5/Forcing/forcing2_gmst.csv",sep=";")
yearIn<-Ftab$Year
## REFERENCE? for the forcing data
```



```{r Effective radiative forcing from AOD/CO2}

# define reference period
# ind.ref<-which((Ftab$Year>1800)&(Ftab$Year<1900))

#FRef<-apply(Ftab[ind.ref,],2,mean)
FRef<-apply(Ftab,2,mean)

dRCO2<-5.35*log(Ftab$CO2/285) #Myrhe et al.
dRTSI<-Ftab$Solar-1365 # unclear reference - Steinhilber?
dRVolc<-Ftab$Volcanic*(-25)  # evolv2k, get forcing by multiplication with calibration factor:
# Following Gregory et al. 2016 Clim Dyn -- 1 AOD= -18+-0.1 W/m2
# 1 AOD = -23 1/m2  Andersson et al. : https://dx.doi.org/10.1038\%2Fncomms8692
# -25*SAOD (IPCC AR5)

RF<-as.data.frame(cbind(Ftab$Year,dRCO2,dRVolc,dRTSI))
colnames(RF)<-c("Year","dR_CO2","dR_Volc","dR_TSI")

matplot(RF$Year,RF[,2:4],type="n",ylim=c(-10,2),xlab="year CE", ylab="relative radiative forcing [W/m^2]")
matlines(RF$Year,RF$dR_CO2,col="brown")
matlines(RF$Year,RF$dR_Volc,col="black")
matlines(RF$Year,RF$dR_TSI,col="orange")
matlines(RF$Year,RF$dR_CO2+RF$dR_Volc+RF$dR_TSI,col="red",lwd=2)
legend("bottomleft",c("CO2","Volcanic","Solar","Total"),col=c("brown","black","orange","red"),lty=rep(1,4),lwd=c(rep(1,3),2))
```





```{r All forcings, warning=FALSE}
rout<-getebm.0d(seq(-500,2000),Flux=cbind(RF$Year,RF$dR_Volc+RF$dR_TSI+RF$dR_CO2))
plot(rout$tsout,xlab="CE",ylab="GMT [C]")
```
```{r Run the EBM with different combinations of solar, warning=FALSE}

rout<-getebm.0d(seq(-500,2000),Flux=cbind(RF$Year,RF$dR_TSI+RF$dR_CO2))
plot(rout$tsout,xlab="CE",ylab="GMT [C]")

# rout<-getebm.0d(seq(-500,2000),Flux=cbind(RF$Year,RF$dR_CO2),plot=FALSE)
# plot(rout$tsout,xlab="CE",ylab="GMT [C]")
# rout<-getebm.0d(seq(-500,2000),Flux=cbind(RF$Year,RF$dR_TSI),plot=FALSE)
# plot(rout$tsout,xlab="CE",ylab="GMT [C]")
# rout<-getebm.0d(seq(-500,2000),Flux=cbind(RF$Year,RF$dR_Volc),plot=FALSE)
# plot(rout$tsout,xlab="CE",ylab="GMT [C]")
# rout<-getebm.0d(seq(-500,2000),Flux=cbind(RF$Year,RF$dR_Volc+RF$dR_TSI),plot=FALSE)
# plot(rout$tsout,xlab="CE",ylab="GMT [C]")

# as we are interested in the multidecadal timescale, adjust the heat capacity of the EBM by modifying the mid-layer depth of the ocean
heatcap.inv(tau=100)$h # effective water column depth
parms<-list(f=0.7,h=1000,S=1365,eps=0.6,alpha=0.3)
rout_volc<-getebm.0d(Flux=cbind(RF$Year,RF$dR_Volc),seq(-500,2000),interp="linear",parms=parms,plot=FALSE)
plot(rout_volc$tsout,xlab="CE",ylab="GMT")

rout_all<-getebm.0d(Flux=cbind(RF$Year,RF$dR_Volc+RF$dR_TSI+RF$dR_CO2),seq(-500,2000),interp="linear",parms=parms,plot=FALSE)


rout_novolc<-getebm.0d(Flux=cbind(RF$Year,RF$dR_TSI+RF$dR_CO2),seq(-500,2000),interp="linear",parms=parms,plot=FALSE)

rout_solar<-getebm.0d(Flux=cbind(RF$Year,RF$dR_TSI),seq(-500,2000),interp="linear",parms=parms,plot=FALSE)

rout_co2<-getebm.0d(Flux=cbind(RF$Year,RF$dR_CO2),seq(-500,2000),interp="linear",parms=parms,plot=FALSE)
```

```{r Plot EBM results, warning=FALSE}
plot(rout_all$tsout,xlab="CE",ylab="GMT [C]",axes=FALSE,xlim=c(0,1850))
axis(1,at=seq(0,2000,by=100))
axis(2)
lines(rout_novolc$tsout,col="blue")
lines(rout_solar$tsout,col="orange")
lines(rout_co2$tsout,col="red")
grid()
legend("bottomleft",text.col=c("black","blue","orange","red"),c("all","solar+CO2","solar","CO2"))

```



############  FURTHER ANALYSIS IDEAS
Idea: Split the volcanic history into parts and generate a "high volcanic frequency" and a "low volcanic frequency" ensemble of forcing surrogates, that are each as long as the entire time series. 
```{r Generating surrogates for the volcanic forcing}


##1 Volcanic frequency - split to get the high/low versions
#thresh<-0.005 # for AOD
#lowyears<-yearIn[which(volc>thresh)] #for AOD
volc<-RF$dR_Volc
thresh<- -0.115 # in W/m^2,


plot(yearIn,abs(volc),log="y")
abline(h=abs(thresh),col="red")
lowyears<-yearIn[which(volc>thresh)] #for AOD
hist(diff(lowyears))

# find those for which diff(lowyears) is more than 3 years to outline segments for permutation
segpoints<-lowyears[(diff(lowyears)>3)]


plot(yearIn[1:segpoints[5]],volc[1:segpoints[5]])
abline(v=segpoints)

#cumsum(yearIn%in%segpoints) # neat trick to split segments!
SP<-split(RF,cumsum(yearIn%in%segpoints)) # list of data frames with forcing


# get the maximum volcanic forcing (max for AOD, min for rel rad. Forcing)
volcmax<-(sapply(SP,function(x){min(x[['dR_Volc']])}))
ranking<-sort(volcmax,index.return=TRUE,decreasing=TRUE)$ix # decreasing for AOD

# find cutoff point to split high/low magnitude eruptions
plot(volcmax[ranking],xlab="ranked segment",ylab="max. rad. forcing W/m^2")
abline(v=length(SP)/2)
# weak positive correlation between log max volc and duration of the interval 
plot(sapply(SP,function(x){log(abs(min(x$dR_Volc)))}),sapply(SP,function(x){length(x$dR_Volc)}))

## Ranked training sets to draw from - construct a high and a low magnitude version
SP.low<-(SP[ranking[1:(length(SP)/2)]])
SP.high<-(SP[ranking[(length(SP)/2+1):(length(SP))]])

```



```{r Construct randomized forcing as Alternative - Example}
pm<-drawfrom(SP,2000,plot=FALSE)
pm2<-drawfrom(SP.low,2000,plot=TRUE,repzero = 1)
pm3<-drawfrom(SP,2000,plot=TRUE,repzero=10,fixvarout = TRUE,targetvarout = var(RF$dR_Volc)+0.01)
pm3<-drawfrom(SP = SP,2000,plot=TRUE,repzero=1,fixvarout = TRUE,targetvarout = 0.8)
# dQ<-rnorm(n = length(RF$dR_Volc),sd = sqrt(1-var(RF$dR_Volc)))
# > var(RF$dR_Volc+dQ)

plot(pm$dR_Volc,type="l")
lines(Ftab$dR_Volc,col="red")
lines(pm2$dR_Volc,col="green")
lines(pm3$dR_Volc,col="blue")
```
## get spectra
```{r Fewer high-magnitude eruptions}
# use sparse volcanic history pm2
parms<-list(f=0.7,h=1000,S=1365,eps=0.6,alpha=0.3)
rout_few<-getebm.0d(Flux=cbind(pm2$Year,pm2$dR_Volc),seq(-500,2000),interp="linear",parms=parms,plot=FALSE)

volccol<-"black"
tempcol<-"blue"

plot(rout_few$tsout,xlab="CE",ylab="GMT",col=tempcol,axes=FALSE)
axis(1)
axis(2,col=tempcol,col.axis=tempcol)
par(new=TRUE)
plot(pm2$Year,pm2$dR_Volc,col="red",type="l",ylim=rev(range(pm2$dR_Volc)),axes=FALSE,xlab="",ylab="")
axis(4,col.axis="red",col="red")

```

```{r set up of the experiment, warning=FALSE}

nmc<-1000
parms<-list(f=0.7,h=70,S=1365,eps=0.6,alpha=0.3)
FF<-c(seq(0,10,by=1),seq(12,20,by=2),c(25,30,40,50))
# FF<-c(seq(0,20,by=1),c(30,40,50,75,100))
FR<-rep(NA,length(FF))
Ftime<-seq(0,1000)
inttime<-seq(-500,1000)
tsc.in<-c(30,200)

RES.md<-as.data.frame(matrix(NA,ncol = nmc,nrow = length(FR)));
colnames(RES.md)<-paste("iter",seq(1,nmc))
row.names(RES.md)<-FF
RES.all<-FORC.all<-RES.md

exmpl<-vector(mode="list",length=length(FR))

getresponse<-function(VolcH,Ftime,inttime,parms){
  tmp<-getebm.0d(Flux=cbind(Ftime,VolcH),inttime,interp="linear",parms=parms,plot=FALSE)
  out<-coredata(tmp$tsout)
  rm(tmp)
  return(out)
}


```

```{r run the model}
##
for (i in 1:length(FF)){
  print(i)
  repzero<-FF[i]
  FMat<-replicate(nmc,drawfrom(SP,max(inttime),plot=FALSE,repzero=repzero,fixvarout = TRUE,targetvarout = var(RF$dR_Volc)+0.001)$dR_Volc)
  # if this is changed from SP to anything else, also replace the targetvarout variance for the TOA flux Q variance
  FORC.all[i,]<-apply(FMat,2,var)
  #LS.f<-apply(FMat,2,function(x,tsc.in){tsc_dep_var(zoo(x,order.by=Ftime),tsc.in)},tsc.in=tsc.in)
  
  exmpl[[i]]<-getebm.0d(Flux=cbind(Ftime,FMat[,1]),inttime,interp="linear",parms=parms,plot=FALSE)
  
  RMat<-apply(FMat,2,getresponse,Ftime=Ftime,inttime=inttime,parms=parms)
  
  LS<-apply(RMat,2,function(x,tsc.in){tsc_dep_var(zoo(x,order.by=Ftime),tsc.in)},tsc.in=tsc.in)
  
  
  # tot.var<-sapply(LS,function(x){x$var.tot})
  mdec.var<-sapply(LS,function(x){x$var.tsc})
  tot.var<-(apply(RMat,2,var)) # total variance
  RES.md[i,]<-mdec.var
  RES.all[i,]<-tot.var
  FR[i]<-mean(mdec.var)/mean(tot.var)
}



```



```{r}


i=1
j=7

cis<-varrat_ci(FR,rep(nmc*LS[[1]]$dof,length(FR)),rep(nmc*LS[[1]]$dof,length(FR)))
pdf(file=paste(outputdir, "multidec_ratio_volcforcing.pdf",sep=""),width=6,height=4)
  par(mar=c(3,3,1,1),cex=1.15,lwd=2)
  plot(FF,FR,xlab="",ylab="",ylim=range(cis),pch=20,col="black",bg="black",axes=FALSE,type="n")
arrows(x0 = FF,x1 = FF,y0=cis[,1],y1=cis[,2],length = 0.01,code = 3,angle=90)
points(FF,FR,pch=20,col="black",bg="black")
points(FF[c(i,j)],FR[c(i,j)],pch=20,col="indianred",bg="indianred")
# what is the average eruption epoch length?
axis(1,at=axTicks(1),labels=FALSE)
mtext(side=1,at=axTicks(1),text=axTicks(1),line=0.5)
axis(2,at=axTicks(2),labels=FALSE)
mtext(side=2,at=axTicks(2),text=axTicks(2),line=0.5)
mtext(side=1,font=2,line=2,"relative number of zero-eruption epochs")
mtext(side=2,font=2,line=1.5,expression(bold(paste(group("",{sigma[30-200]^2}/sigma[tot]^2,"")))))
box()
dev.off()
```
```{r}
plot(FF,apply(RES.md,1,mean),ylab="var(30-200a)",xlab="expected number of zero-eruption epochs")
plot(FF, apply(RES.all,1,mean),xlab="expected number of zero-eruption epochs",ylab="var(total)")

```





```{r Time series illustrations}

# plotxx<-function(ts1,ts2,col1="black",col2="blue"){
# 
# plot(ts1,col=col1,xlim=range(index(ts1),index(ts2)),ylim=range(ts1,ts2))
# lines(ts2,col=col2)
# }
# 
# plotxx(exmpl[[1]]$tsout,exmpl[[3]]$tsout)

## plot forcing and response

tsx2<-exmpl[[j]]$tsout

tsy2<-zoo(exmpl[[j]]$forcing.ts,order.by=index(exmpl[[j]]$tsout))

for (k in c(1,7,20)){
pdf(file=paste(outputdir,"ts_",k,".pdf",sep=""),width=6,height=4)
  yax2=FALSE
par(mar=c(4,4,1,4),cex=1.15,lwd=2,cex.lab=1.15)
tsx1<-exmpl[[k]]$tsout
tsy1<-zoo(exmpl[[k]]$forcing.ts,order.by=index(exmpl[[k]]$tsout))
if (k>6) yax2=seq(-3,0)
plotxy(tsx1,tsy1,col1="black",ylimy1=c(-1,0.2))
dev.off()
}

# plotxy(tsx2,tsy2,col1="black")
# in principle, the variance of the forcing is to a value slightly higher than in the original volcanic forcing, but drawing only one realization for each recurrence strengths, the effective forcing strength varies slightly.
#plot(sapply(exmpl,function(x){var(x$forcing.ts)}),sapply(exmpl,function(x){var(x$tsout)}))

```
** TO BE DONE**
## more systematically: get variances of the forcing ts and variance of the responses 
## check that the forcing variance is always the same
## get the response for a systematic range
## get quantiles and dual plots for the examples
## look at the spectra of forcing and response
## read up on spectral gain
## plot decadal vs. total variance <----------------------------- I think that's the key point


#boxplot(log10(yonkers), log10(stamford),names = c("Yonkers", "Stamford"),main = "New York Ozone Levels",ylab = "Log10 Ozone Levels (ppb)",notch = TRUE)
#rout_few<-getebm.0d(Flux=cbind(pm2$Year,VolcH),seq(-500,2000),interp="linear",parms=parms,plot=FALSE)

